Turbulence and Particle Acceleration in Giant Radio Halos: the Origin of Seed Electrons

### 2016-04-21 paper.tex
intro polished

There have been two competing models proposed to explain RHs.  The
radio emitting electrons in the ``hadronic model'' are produced in
inelastic (hadronic) CR proton interactions with protons of the
ambient thermal ICM, which generates pions that eventually decay into
electrons and positrons, depending of the charge of the initial pion
(\citealp{1980ApJ...239L..93D,1999APh....12..169B,2001ApJ...562..233M,
  2004A&A...413...17P,2008MNRAS.385.1211P,ensslin11}). CR protons and
heavier nuclei may have been accelerated and injected into the ICM by
structure formation shocks, active galactic nuclei and galactic
winds. However, the strong bimodality that separates X-ray luminous
clusters into radio-active and radio-quite clusters (and requires a
fast switch on/off mechanism of the RH emission) and the very extended
RH emission at low frequencies in Coma (352~MHz) represent a major
challenge to this model class \citep{brunetti12,2014MNRAS.438..124Z}.

@@

However, there is a salient piece missing in the turbulent
reacceleration model. It relies heavily on the assumption of an
abundant, volume-filling population of seed suprathermal electrons;
direct Fermi II acceleration from the thermal pool is precluded by
strong Coulomb losses
\citep{2008ApJ...682..175P,2012ApJ...759..113C}. These seeds are
presumed to be either fossil CR electrons (CRes) accelerated by
diffusive shock acceleration (DSA) during structure formation
\citep{1999ApJ...520..529S}, or secondaries injected by hadronic
interaction of CR protons (CRps) with thermal protons
\citep{brunetti11}. While analytic estimates have been made, there has
been no ab initio demonstration that structure formation can lead to
the required abundance of seed electrons with the correct spatial and
spectral characteristics. This is a non-trivial requirement: Coulomb
cooling in dense cluster cores is severe, and DSA fossil electrons may
not survive. On the other hand, for secondaries to constitute the seed
population, the CRp population required in the best-studied case of
the Coma cluster must have a very broad and flat (or even slightly
inverted) spatial profile \citep{brunetti12}, in contrast with the
thermal plasma whose energy density declines steeply with radius. In
Fig.~\ref{fig:Edens} we show that such a distribution is not predicted
by cosmological simulations \cite[see
  also][]{pinzke10,2014MNRAS.439.2662V}. If CRps are predominantly
advected with the cluster plasma, their distribution will be peaked
towards the cluster center and shows a similar characteristics as the
thermal plasma. As a consequence, the distribution of secondary
electrons and the resulting radio synchrotron emission is also peaked
since the hadronic reaction is a two-body scattering process (see
Fig.~\ref{fig:Edens}). Hence, the simulated emission falls short
of the observed extended and flat radio profile of the Coma cluster.

@@

Indeed, arriving at a seed population with the required
characteristics is highly constraining, and has the potential to teach
us much about the origin of CRps/CRes in clusters.  In this work, we
use our hydrodynamical zoom simulations of galaxy clusters in a
cosmological setting to follow the distribution functions of seed CRp
and CRe population and integrate the Fokker-Planck equation of CR
transport along Lagrangian particle trajectories. We resolve injection
at structure formation shocks, account for various loss processes of
CRs, and -- most importantly -- model second-order Fermi acceleration
by CR interactions with magnetised turbulence. However, we assume a
simplified and stationary model for magnetic fields and turbulence. We
do not account for the time-varying energy density in compressible
waves, which are thought to be necessary for the acceleration process
\citep{brunetti07,brunetti11}, as the cluster merger proceeds.  So our
approach is orthogonal (and complementary) to e.g.,
\citealp{2015ApJ...800...60M}. This enables us to vary parameters
associated with the unknown continuation of magneto-hydrodynamical
turbulence below the MHD scale and the overall amplitude of
compressible waves.

@@

To pursue these three possibilities further, we employ cosmological
simulations of CRs in clusters. In tandem with new insights from our
recent work on DSA generated fossil electrons \citep{pinzke13}, we
generate the first quantitative calculation of primary and secondary
seed electrons and compare to observations in the Coma cluster. For
the most part, we adopt parameters that are used in previous work: for
details, see \citep[][ turbulent spectrum]{brunetti07,brunetti11},
\citep[][ injection scale of turbulens]{2015ApJ...800...60M},
\citep[][ energy density of
  turbulence]{2009ApJ...705.1129L,2010ApJ...725.1452S,2011A&A...529A..17V},
for CR acceleration in shocks, \citep[][ in particular $K_{\rmn{ep}}$,
  acceleration efficiency]{pinzke13}, and for cosmological
simulations, \citep[][ i.e., shock history, temperature profile,
  density profile, magnetic field profile]{pinzke10}. All these
parameters are uncertain, and could potentially change our
predictions. We refer the reader to section~\ref{sect:param_comp} for
a minor exploration of parameter space, and leave the full exploration
of parameter space and the interaction between free parameters to
future work.

@@

The transport of relativistic electrons and protons in the ICM is a
complex process that depends both on the details of the thermal
component (gas density, temperature, and pressure) as well as
non-thermal component (turbulence, magnetic fields, fossil CRs). We
use high resolution galaxy cluster simulations to derive the thermal
and fossil CR properties (shock accelerated primary CRes and CRps, as
well as secondary CRes produced in p-p collisions)
\citep{2007MNRAS.378..385P,pfrommer08,pinzke10,pinzke13}.

>>>>

There have been two competing models proposed to explain RHs.  The
radio emitting electrons in the ``hadronic model'' are produced in
inelastic (hadronic) CR proton interactions with protons of the
ambient thermal ICM, which generates pions that eventually decay into
electrons and positrons, depending of the charge of the initial pion
(\citealp{1980ApJ...239L..93D,1999APh....12..169B,2001ApJ...562..233M,
  2004A&A...413...17P,2008MNRAS.385.1211P,ensslin11}). CR protons and
heavier nuclei may have been accelerated and injected into the ICM by
structure formation shocks, active galactic nuclei and galactic
winds. However, the strong bimodality that separates X-ray luminous
clusters into radio-active and radio-quite clusters (requiring a
fast switch on/off mechanism of the RH emission) and the very extended
RH emission at low frequencies in Coma (352~MHz) represent a major
challenge to this model class \citep{brunetti12,2014MNRAS.438..124Z}.

@@

However, there is a salient piece missing in the turbulent
reacceleration model. It relies heavily on the assumption of an
abundant, volume-filling population of seed suprathermal electrons;
direct Fermi II acceleration from the thermal pool is precluded by
strong Coulomb losses
\citep{2008ApJ...682..175P,2012ApJ...759..113C}. These seeds are
presumed to be either fossil CR electrons (CRes) accelerated by
diffusive shock acceleration (DSA) during structure formation
\citep{1999ApJ...520..529S}, or secondaries injected by hadronic
interaction of CR protons (CRps) with thermal protons
\citep{brunetti11}.

While analytic estimates have been made, there has been no ab initio
demonstration that structure formation can lead to the required
abundance of seed electrons with the correct spatial and spectral
characteristics. This is a non-trivial requirement: Coulomb cooling in
dense cluster cores is severe, and DSA fossil electrons may not
survive. On the other hand, for secondaries to constitute the seed
population, the CRp population required in the best-studied case of
the Coma cluster must have a very broad and flat (or even slightly
inverted) spatial profile \citep{brunetti12}, in contrast with the
thermal plasma whose energy density declines steeply with radius. In
Fig.~\ref{fig:Edens} we show that such a distribution is not predicted
by cosmological simulations \cite[see
  also][]{pinzke10,2014MNRAS.439.2662V}. If CRps are predominantly
advected with the cluster plasma, their distribution will be peaked
towards the cluster center and shows a similar characteristics as the
thermal plasma. As a consequence, the distribution of secondary
electrons and the resulting radio synchrotron emission is also peaked
since the hadronic reaction is a two-body scattering process. Hence,
the simulated emission falls short of the observed extended and flat
radio profile of the Coma cluster.

@@

Indeed, arriving at a seed population with the required
characteristics is highly constraining, and has the potential to teach
us much about the origin of CRps/CRes in clusters.  In this work, we
use our hydrodynamical zoom simulations of galaxy clusters in a
cosmological setting to follow the distribution functions of seed
populations for CRps and CRes, and integrate the Fokker-Planck
equation of CR transport along Lagrangian particle trajectories. We
model diffusive shock acceleration at structure formation shocks,
account for various loss processes of CRs, and -- most importantly --
model second-order Fermi acceleration by CR interactions with
magnetised turbulence.

However, we assume a simplified and stationary model for magnetic
fields and turbulence. We do not account for the time-varying energy
density in compressible waves, which are thought to be necessary for
the acceleration process \citep{brunetti07,brunetti11}, as the cluster
merger proceeds.  So our approach is orthogonal (and complementary) to
e.g., simulations of \citealp{2015ApJ...800...60M} that focus on the
time-dependent compressible turbulence while adopting a simplified
treatment of CR that does not follow the transport of the CR
distribution function. Our approach of parametrizing turbulence
enables us to vary parameters associated with the unknown continuation
of magneto-hydrodynamical turbulence below the MHD scale and the
overall amplitude of compressible waves (that can in principle vary
depending on the details of a particular cluster merger).

@@

To pursue these three possibilities further, we employ cosmological
simulations of CRs in clusters. In tandem with new insights from our
recent work on DSA generated fossil electrons \citep{pinzke13}, we
generate the first quantitative calculation of primary and secondary
seed electrons and compare to observations in the Coma cluster. For
the most part, we adopt parameters that are used in previous work,
including Fermi-II acceleration with models of turbulent spectra
\citep{brunetti07,brunetti11}, injection scale of turbulence
\citep{2015ApJ...800...60M}, the energy density of turbulence
\citep{2009ApJ...705.1129L,2010ApJ...725.1452S,2011A&A...529A..17V},
for CR acceleration in shocks \citep[][in particular $K_{\rmn{ep}}$
  and the acceleration efficiency]{pinzke13}, and for cosmological
simulations with CR physics \citep[][i.e., shock history,
  temperature profile, density profile, magnetic field
  profile]{pinzke10}. All these parameters are uncertain, and could
potentially change our predictions. We refer the reader to
section~\ref{sect:param_comp} for a first exploration of parameter
space, however we leave the detailed exploration of parameter space and
the interaction between free parameters to future work.

@@

The transport of relativistic electrons and protons across cosmic time
into galaxy clusters is a complex problem that depends on the velocity
field of the gas (and its thermodynamic properties such as density,
temperature, and pressure) as well as non-thermal processes
(turbulence, magnetic fields, fossil CRs). We use high resolution
galaxy cluster simulations to derive the thermal and fossil CR
properties (shock accelerated primary CRes and CRps, as well as
secondary CRes produced in p-p collisions)
\citep{2007MNRAS.378..385P,pfrommer08,pinzke10,pinzke13}.
